Human Factors Influencing Visual Statistical Inference

### 2013-06-07 sociology_paper.tex
finished with intro

\usepackage{color}
\newcommand{\hh}[1]{{\color{magenta} #1}}

@@

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examine these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect. There is evidence that a learning effect exists that observers get better and faster with repeated evaluations which suggests a potential for using the lineup protocol for training people to read statistical graphics. The placement of actual data plot in the lineups does not affect the inference.

@@

In visual inference, the test statistic is a plot of the observed data. To create a lineup, this plot, called actual data plot, is placed in a layout of null plots. The null plots  are generated from the model specified by a null hypothesis, essentially describing what the plot might look like if the data had no structure. An observer is asked to evaluate the lineup. If the  actual data plot is detected by the observer, the null hypothesis is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness. Combining the choices of multiple observers provides more stability in the estimation of significance.

\subsection{Visual inference example}

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which of these towers is the most extreme? Which side is higher? In an experiment 12 of 72 observers picked plot \# $3^2+4$ as the most extreme plot from the lineup, based on the highest amount of asymmetry (36\%), a trend (26\%) or `outliers' (16\%). 
This gives us a visual $p$-value of 0.00023, indicating that we have enough evidence to reject the null hypothesis (in another four repetitions we found similar results, giving an overall $p$-value of less that $10^{-17}$ based on 62 data picks out of 341 evaluation). 
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots: in this example, we are investigating, whether the results from the 2012 US presidential election is consistent with polling results. As a null hypothesis we  assume that  the election results are consistent with polling results. The polling results give us a null model to generate data. Each poll gives us a margin between candidates. This is what we use as the mean of a normal distribution. In the absence of information of error for a poll, we will assume a standard deviation of 2.5 percentage points for all of the polls. While this is a severe simplification, the resulting margin of error of $\pm 5$ percentage points is realistic, and could be easily replaced if additional information on a poll became available. A null data set is then generate as a set of draws from normal distributions based on the latest poll results. Each panel in figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes}: each state is represented by a rectangle. The width of a rectangle is determined by the difference between (polling) results, its height is given by the number of electoral votes, and fill color shows party affiliation (blue for Democrat, and red for Republican). Rectangles are placed on top of each other, separated by affiliation and sorted by margin between polls. A party needs 270 electoral votes or more to win the presidential election.

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{tower.pdf} 
   \caption{ \label{fig:tower} Which of these towers is the most extreme? Which side is higher?}
\end{figure}

@@

If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand, if the null hypothesis is not true, the observed data should be different from the rest of the plots in the lineup and observers should be able to pick the data plot out from the rest.

A lineup can be evaluated by a single person or multiple observers. Based on the observer feedback the $p$-value is computed \citep{majumder:2013} and finally a decision is made about the hypothesis being tested.

@@

\subsection{Observer task} Observer who evaluates a lineup do not necessarily be aware of the data that constitue a lineup. In response to a specific question the observer is to make a choice based on what he or she sees in the lineup. The choice could be right or wrong depending on whether the data plot is selected or not. Under null hypothesis, all the null plots look similar to what the observed data plot appears to be. This would make it harder to detect the observed data plot in the lineup. It is not expected that an observer would be able to detect the data plot in this scenario. But since there are limited number of plots in a lineup which is 20 in this case, there is a 1/20 chance that the observer would pick the actual plot. This proportion is associated with the type-I error probability. On the other hand if null hypothesis is not true, the observed plot would be different than the null plots making it easier to be distinguished. When multiple observers evaluate a lineup, the proportion of correct response can be used to estimate the power of the visual test \citep{majumder:2013}.

For a lineup with fixed difficulty, the proportion correct responses may vary for different observers. If an observer evaluates multiple lineups, the proportion of correct evaluations by the observer could be used as a measure of performance by an observer. The another way of measuring the performance of the observer could be the time taken by an observer to evaluate a lineup. Since performance would also depend on the lineup difficulty it is essential to adjust for lineup difficulty before estimating the performance of the observer.

Some of the factors that may affect the performance of the observer are discussed in Section \ref{sec:factor_performance}. In this paper we focus on estimating the learning effect of the observer in terms of both proportion correct and time taken to evaluate a lineup through the attempts made during the evaluation of multiple plots. We also estimate the location effect of the observed data plot in a lineup. The other factors are controlled in simulation experiments so that learning effect and location effect can be estimated.

This paper is organized as follows; Section \ref{sec:factor_performance} describes some of the factors that may affect the performance of the observer. Section \ref{sec:exp_design} describes the design of the simulation experiments and methods to estimate the learning trend of the observer and location effect of the actual plot in a lineup. Section \ref{sec:result_socio} presents the analysis of the experimental data and the results.

\subsection{Experiments pooled}

>>>>

\usepackage{color}
\newcommand{\hh}[1]{{\color{magenta} #1}}
\newcommand{\dc}[1]{{\color{green} #1}}

@@

\begin {abstract}  
Visual statistical inference is a way to determine significance of patterns found while exploring data. It is dependent on the evaluation of a lineup, of a data plot among a sample of null plots, by human observers. Each individual is different in their cognitive psychology and judiciousness, which can affect the visual inference. The usual way to estimate the effectiveness of a statistical test is its power. The estimate of power of a lineup can be controlled by combining evaluations from multiple observers. Factors that may also affect the power of visual inference are the observers' demographics, visual skills, and experience, the sample of null plots taken from the null distribution, the position of the data plot in the lineup, and the signal strength in the data. This paper examine these factors. Results from multiple visual inference studies using Amazon's Mechanical Turk are examined to provide an assessment of these. The experiments suggest that individual skills vary substantially,  but demographics do not have a huge effect. There is evidence that a learning effect exists that observers get better and faster with repeated evaluations which suggests a potential for using the lineup protocol for training people to read statistical graphics. The placement in the lineup does not affect the inference.

@@

In visual inference, the test statistic is a plot of the observed data. To create a lineup, this plot, called the actual data plot, is placed in a layout of null plots. The null plots  are generated from the model specified by a null hypothesis, essentially describing what the plot might look like if the data had no structure. An observer is asked to evaluate the lineup. If the  actual data plot is detected by the observer, the null hypothesis is rejected. This means that the structure in the actual data plot has significant structure, a pattern that is not simply due to randomness. Combining the choices of multiple observers provides more stability in the estimation of significance.

%\subsection{Visual inference example}

Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which one of the 20 plots is the most different from the others? When asked this question, 12 of 72 observers picked the same plot (with number equal to the result of $3^2+4$), with reasons given being 'asymmetry' (36\%), 'trend' (26\%) or `outliers' (16\%).  The corresponding $p$-value is 0.00023, indicating sufficient evidence to reject the null hypothesis.  What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots. This example investigates the results from the 2012 US presidential election in comparison to the poll results just prior to the election. (Although this example is more simplistic than most of the tests conducted to date, it will serve the purpose of illustrating the lineup protocol.) The data is looking at the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes} where each state in the union is represented by a rectangle. This difference is plotted horizontally, and the height of each box corresponds to the state's electoral votes. Color indicates party affiliation.  The null hypothesis is ``that the election results were consistent with the polls''. The polling results provides the null model from which data is simulated. Because each poll has a margin of error, this is used to simulate different scenarios that might have resulted on election day, if the polls were on target. A null data set is then generate as a set of draws from normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, qpproximating a margin of error of 5\%. These samples are plotted as electoral buildings, and the plot made with the results from the election is placed randomly among them in a lineup of size $5\times 4$. If the null hypothesis is true then the actual data plot should look just like any of the other plots, and not be identified by an observer. %A party needs 270 electoral votes or more to win the presidential election.

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{tower.pdf} 
   \caption{ \label{fig:tower} \dc{Which one of the plots is the most from the others? Maybe this question will have people picking 4, because it is the only one with red higher.}}
\end{figure}

@@

%If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand, if the null hypothesis is not true, the observed data should be different from the rest of the plots in the lineup and observers should be able to pick the data plot out from the rest.

A lineup can be evaluated by a single person or multiple observers. A \dc{multinomial} distribution is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis.  Observers should not be aware of the data that constitutes a lineup, and should not have seen the actual data plot before seeing the lineup. This is the reason that in the election example, above, the scenario was explained after the lineup question, in the text.

The question that is asked of the observer should be as general as possible, effectively asking the observer to pick the plot that is different, and allowing them to provide their reasons for seeing their pick as different. In some of the studies, the ones described in  \cite{majumder:2013} very specific questions were asked, because the experiments were being conducted to compare results from the lineup protocol with those of conventional tests. In those experiments, structure in the data was strictly controlled in the simulation process, which allowed for specific questions to be asked. In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is most difference between the two heights. Most observers got this, but it is possible that some observers might pick plot 4, where the red tower is slightly above the blue as the most different because it is the only plot with this feature. So a better question may have been ``Which plot shows the biggest height difference between the two towers?'' except that this tailors the inference to a specific feature which does not match the null hypothesis of interest.

@@

%\subsection{Observer task}

For any evaluation the observer may or may not identify the actual data plot. Under the null hypothesis, the actual data plot should look similar to the null plots making it harder to detect. It is not expected that an observer would be able to detect the data plot in this scenario. But since there are limited number of plots in a lineup, which is 20 in the election example, there is a 1/20 chance that the observer would pick the actual plot. This proportion is associated with the Type~I error of the test. On the other hand, if null hypothesis is not true, the observed plot should look different from the null plots, making it easier to be detected. This is the definition of the power of the test. When multiple observers evaluate a lineup, the proportion of correct response can be used to estimate the power. The ability of individual observers can vary, and examining this is the purpose of this paper.

%For a lineup with fixed difficulty, the proportion correct responses may vary for different observers. If an observer evaluates multiple lineups, the proportion of correct evaluations by the observer could be used as a measure of performance by an observer. The another way of measuring the performance of the observer could be the time taken by an observer to evaluate a lineup. Since performance would also depend on the lineup difficulty it is essential to adjust for lineup difficulty before estimating the performance of the observer.

%Some of the factors that may affect the performance of the observer are discussed in Section \ref{sec:factor_performance}. In this paper we focus on estimating the learning effect of the observer in terms of both proportion correct and time taken to evaluate a lineup through the attempts made during the evaluation of multiple plots. We also estimate the location effect of the observed data plot in a lineup. The other factors are controlled in simulation experiments so that learning effect and location effect can be estimated.

There have been \dc{ten} experiments conducted using Amazon's Mechanical Turk~\cite{turk} that are being used to evaluate the the effect of the observer. Table \ref{tbl:visual_stat} summarizes these. Experiments XXX are used to evaluate the effects of demographic factors whethere there is a learning trend. The design of experiment XXX enabled effect of position in the lineup, and the sample of nulls, to be evaluated. Section \ref{sec:factor_performance} describes the factors that may affect the performance of the observer. Section \ref{sec:exp_design} describes the methods used to assess the effects, and Section \ref{sec:result_socio} describes the results.

%\subsection{Experiments pooled}
